Assessing \Htwo{} and \COtwo{} Network Strategies in Europe

### 2023-12-14 report/submissions/journal/paper/paper.tex
follow up
paper: address todos in intro plots: minor changes

\title{Assessing \Htwo{} and \COtwo{} Network Strategies in Europe}
% \title{Synthetic fuels in Europe: Transport Hydrogen to Carbon, or Carbon to Hydrogen?}

@@

The transition to a carbon-neutral European economy is a pressing challenge that demands coordinated action across various energy sectors. While management of both \COtwo{} and \Htwo{}, also refered to as carbon and hydrogen, is considered a critical component of this transition, a gap exists in understanding how new hydrogen infrastructure effectively interacts with carbon management technologies, including carbon capture, transport, storage, utilization, and sequestration. Hydrogen offers an efficient way to transport and store energy over long distances and timeframes, and can be produced from electrolysis or steam methane reforming (SMR), for example. Carbon can be stored in geological formations to permanently extract it from the atmosphere, a process known as carbon sequestration (CS). Additionally, carbon can be combined with hydrogen to produce carbonaceous fuels needed for aviation (synthetic kerosene from Fischer-Tropsch synthesis), shipping (synthetic methanol) and industrial feedstocks (e.g., synthetic methane). From hereon, we jointly refer to the production of these carbonaceous fuels as Carbon Utilization (CU). Carbon can be effectively captured from industrial processes and the combustion of biomass, fossil fuels, or synthetic carbonaceous fuels through point-source carbon capture (CC) techniques, or harvested from the atmosphere using direct air capture (DAC). In combination, \COtwo{} and \Htwo{} networks are crucial to enable these carbon management strategies for net-zero economies.
% TODO: integrate definition of Carbon Management (CM) into last sentence.
% Cx: CU, CS, CT, CC, Cstorage --> CO2 + H2 networks are crucial to carry out CM strategies
% Suggestion:

% TODO: check definition of carbonaceous fuels for "missing" (relevant) fuel types

% TODO: boil this down to one sentences and integrate it to the previous paragraph or move it two the literature part.
% Suggestion: We argue that it is crucial to consider CC, CU and CS as separate technologies, rather than grouping them under the often-used but misleading umbrella term "CCUS", to allow for more accurate optimization and analysis of each function in the carbon management system.
% Suggestion: We argue that it is crucial to consider CC, CU and CS as separate technologies, rather than grouping them under the often-used but misleading umbrella term ``carbon capture, utilization and storage'' (CCUS), to allow for independent optimization and adequate representation of each function in the carbon management system.
It is important to differentiate between the functions of carbon capture, carbon sequestration and carbon utilization. While both academic literature and common language mostly use the term ``carbon capture, utilization and storage'' (CCUS), or similar, to refer to all three functions, we argue that this is misleading. There is inherent benefit in separating these three functions geographically (linked through carbon transport), temporally (linked through carbon storage), and in terms of economic actors performing each function. CC should be performed at the place and with the technology that can capture carbon at the lowest cost (including a potential transport premium). CS should take place where sequestration potentials exist and carbon is available at the lowest cost, and CU where the combination of hydrogen and carbon inputs is available at lowest cost. By employing this separation in our model, we allow for a nuanced optimization and detailed analysis of the carbon management system.

Recently, policymakers and industry in Europe have started developing hydrogen and carbon management strategies~\cite{GermanyDevelopingStrategy2023,CarbonManagementStrategie}, planning infrastructure components~\cite{CONetz}, and committing on the first carbon utilization projects~\cite{EFuelsPilotPlant2022,OrstedAssumesFull,GROUNDBREAKINGEFUELPRODUCTION,DLREfuelsDLR}. With the European Union's goal of achieving climate neutrality by 2050, a wide range of programs, funding models and initiatives have been established in this regard~\cite{eu2023netzero,europeangreendeal,europeaninnovationfund}.
% TODO: mention policy ambition for carbon and hydrogen strategies separately in two sentences.
Collaborative industry initiatives like the European Hydrogen Backbone~\cite{gasforclimateEuropeanHydrogenBackbone2022} or the Hydrogen Infrastructure Map~\cite{H2InfrastructureMap} showcase the potential of hydrogen as a fuel and energy carrier, and some gas pipelines are already repurposed to transport hydrogen~\cite{RohrFreiFuer}. Business models from companies like Tree Energy Solutions~\cite{TESHydrogenLife2023}, Carbfix~\cite{WeTurnCO2}, and Equinor~\cite{adomaitisEquinorRWEBuild2023} advertise carbon management hubs that provide green hydrogen, methane, and synfuels on the one hand and offer \COtwo{} offtake on the other hand. The Northern Lights project in Norway~\cite{NorthernLightsWhat} is planning with a transport and sequestration capacity of 1.5~Mt \COtwo{} per year to be operative in 2024, expanding to a targeted scale of 5~Mt per year of sequestration by 2030.
% TODO: frame next two sentences to be more market oriented
% TODO: rather cite europeancommissionEuropeanCO2Storage
For Europe, the Capture Map~\cite{ToolsGreenTransition} estimates a potential of 1.7~Gt of carbon capture from point sources per year, which represent roughly 50\% of all current emissions. In combination with large sequestration potentials as stated in~\cite{weiProposedGlobalLayout2021}, this highlights the vast potential for carbon capture.
% TODO: optionally find better place for this
To this end, the Clean Air Task Force underlines the importance of a carbon transport system in Europe to facilitate the carbon economy~\cite{lockwoodEuropeanStrategyCarbon}. Considering carbon pipeline as a technological option is a reasonable model assumption since \COtwo{} pipelines are a mature technology, given their wide-spread installations in the United States and Canada for enhanced oil recovery~\cite{righettiSitingCarbonDioxide2017,friedmannNETZEROGEOSPHERICRETURN}.

% TODO: explain why misalignment takes place -> exogenous factors renewable potentials, industrial clusters with h2 demand, co2 point sources, sequestration sites,
However, up to this point, it remains unclear how the two transport systems of hydrogen and carbon may complement or replace each other. Both can bridge the misalignment of sources (electrolysis, SMR, point-source CC, DAC) and sinks (CU, CS, industrial hydrogen demand) for carbon and hydrogen. A hydrogen network can supply regions with geographically fixed hydrogen demand, such as for steelmaking, with hydrogen from regions with the best renewable resources as well as enable CU at the site of point-source CC. On the other hand, the carbon transport system allows for transporting captured carbon to regions with sequestration potentials or high-quality renewable resources, the latter enabling cost-effective CU.
In the literature, the two network approaches and underlying technologies have been discussed in a number of publications, all of which, however, dealt with the isolated effects~\cite{bakkenLinearModelsOptimization2008,morbeeOptimisedDeploymentEuropean2012,stewartFeasibilityEuropeanwideIntegrated2014,oeiModelingCarbonCapture2014,elahiMultiperiodLeastCost2014,burandtDecarbonizingChinaEnergy2019,middletonSimCCSOpensourceTool2020,bjerketvedtOptimalDesignCost2020,weiProposedGlobalLayout2021,damoreOptimalDesignEuropean2021,becattiniCarbonDioxideCapture2022,neumannBenefitsHydrogenNetwork2022}. Such techno-economic models, in contrast for example to Integrated Assessment Models, can account for the spatial distribution of carbon sources and sinks which are crucial to provide a holistic view of the energy system and its technological interactions. A remarkable example was developed by Neumann et al.~\cite{neumannBenefitsHydrogenNetwork2022}, which comprehensively demonstrates the interchangeability of hydrogen and electricity grid expansion in a climate-neutral European energy system at the expense of higher investments. The underlying, highly-resolved model encompasses most relevant sectors, however neglects the option to transport carbon across regions.
The publication by Morbee et al.~\cite{morbeeOptimisedDeploymentEuropean2012} optimizes the topology and capacity of a \COtwo{} network in Europe, but only considers the power sector without co-optimizing renewable deployment. This limited sectoral scope cannot capture important dynamics of carbon management, since it neglects the sectors that will need to handle most \COtwo{} in the future.
A comprehensive example is found in~\cite{becattiniCarbonDioxideCapture2022}, which presents a mixed-integer model to optimize the time-evolution of a \COtwo{} transport system in Switzerland, connecting to a remote sequestration site in Norway. However, this limited spatial scope fails to consider other sequestration sites and co-benefits from connecting the \COtwo{} network to neighboring countries.
Other high-resolution energy system models do not feature detailed CU technologies and \COtwo{} transport. Euro-Calliope, for example, does neither feature carbon nor hydrogen transport, and thus forces generic synthetic fuel production to rely on captive electrolysis and DAC in the same region~\cite{pickeringDiversityOptionsEliminate2022}.

To our knowledge, no study has yet considered the co-optimization and comprehensive assessment of both \COtwo{} and \Htwo{} networks in a fully sector-coupled energy system. However, we argue that such an assessment is strongly needed to identify realms of competition and synergy between hydrogen and carbon management technologies. In this paper, we present a detailed study of the European energy system for 2050, which includes high geographical resolution and a comprehensive representation of carbon management technologies. The study is conducted using the PyPSA-Eur energy system model and encompasses all relevant energy sectors. We investigate the competitive transport dynamics between \COtwo{} and \Htwo{} networks. This is done by contrasting models that feature only one of these networks, neither, or both combined. We also analyze how an energy system with limited annual sequestration potential prioritizes decarbonization and fuel switching in various sectors, and how the construction of carbon and hydrogen networks varies given net-zero or net-negative emissions targets.

We define two \COtwo{} target scenarios:
\begin{itemize}
    \item[] \textit{Net\=/Zero}: aligning with the EU's 2050 emission targets. Limiting carbon sequestration to 200~Mt/a. Also denoted as NZ.
    \item[] \textit{Net\=/Negative}: 10\% net-negative emissions relative to 1990 levels, equaling 460~Mt \COtwo{} annually. Limiting carbon sequestration to 660~Mt/a. Also denoted as NN.
\end{itemize}
Unless otherwise mentioned, our reference is the Net-Zero scenario.

Beyond \COtwo{} targets, we consider four expansion scenarios, referred to as models:
\begin{itemize}
    \item[] \textit{Baseline:} Neither \COtwo{} nor \Htwo{} networks are constructed.
    \item[] \textit{\COgrid{}:} Only the \COtwo{} network is developed.
    \item[] \textit{\Hgrid{}:} Only the \Htwo{} network is developed.
    \item[] \textit{Hybrid:} Both \COtwo{} and \Htwo{} networks are developed.
\end{itemize}

% TODO add two sentences about the used methods for transition

@@

\begin{figure}[ht!]
    \centering
    \includegraphics[width=\linewidth]{comparison/default/figures/90_nodes/cost_bar.png}
    \caption[short]{Total annual system cost subdivided into groups of technologies for the different models of the European energy system with a net-zero emission target. While the \modBase{} has neither a \COtwo{} nor an \Htwo{} network, the \modHybrid{} is allowed to expand both. ``Gas Infrastructure'' combines gas facilities for power and heat production, ``\COtwo{} Infrastructure'' combines transport, storage and sequestration, and ``\Htwo{} Infrastructure'' combines transport and storage. ``Carbon Capture at Point Sources'' combines all technologies with integrated carbon capture, including the cost of the main facility (e.g., CHP unit) and the carbon capture application.}
    \label{fig:cost_bar}
\end{figure}

Total annual system costs in the \modBase{} are highest, totalling \label{}779 billion euros. The models with additional transport options for carbon and hydrogen achieve a more efficient system allocation and lower total system costs, with \label{}3.1\% cost reduction (\label{}754~bn€) in the \modCO{}, and \label{}4.8\% (\label{}741~bn€) in the \modH{}, and \label{}5.2\% (\label{}738~bn€) in the the \modHybrid{}.

@@

% TODO: add interpretative paragraph: no CU where low-price ely, H2 is imported.
% Both networks available to smooth prices; CS draws in CC from adjacent regions; CU everywhere locally, except the CS sites; DAC centered on (very few) regions with optimal RES (and Italy/Greece for methanation)

@@

In transitioning from a net zero to a net emissions reduction target, a continuation of investment trends are perceived. Fig.~\ref{fig:net-negative_cost_bar} shows the net investment changes per technology group for the \carbonmodel, the \Hgrid{} and \modHybrid{}s. The scenario imposes a net carbon removal of 460 Mt/a, with all models incurring a roughly similar net cost increase - 77 billion €/a for the \modCO{}, 81 billion €/a for the \modCO{}, and 78 billion €/a for the \modHybrid{}, balancing the earlier cost advantage of the \modHybrid{}. Both models invest relatively evenly in additional DAC, wind and heat pump systems, with DAC accounting for more than a third of the additional system costs. These remove carbon directly at sequestration sites at the coast, in particular in Portugal and UK, together with heat and power from new heat pumps at site and new solar and wind installation in the vicinity.
More heat pumps, solar and wind installations across all Central and Eastern European countries as well as biogas-to-gas installations compensate for the reduction of fossil gas based heating and powering. Reduced usage of gas in general lower the investments in gas infrastructure in both models.
Fundamental difference in investment strategies occur in the carbon capture sector: In contrast to the \modH{}, the \modHybrid{} continues the strategy of sequestering carbon captured and transported from bio-energetic inputs, expanding the carbon transport routes from Central and Eastern Europe to the near shores. At the same time, carbonaceous fuel production in Central Europe partially moves to Spain and UK, where it uses carbon from new DAC facilities without the installation of a \COtwo{} network. At the same time, solar based electrolysis in France, Nothern Italy and Switzerland expands requiring more solar capacities but less hydrogen imports from the Iberian Peninsula.
The \modH{} on the other hand, expands decentralized CU systems from the Net-Zero scenario, which are now increasingly supplied with carbon from biogas-to-gas facilities, showing up in both ``Bioenergy'' and ``Carbon Capture at Point Sources'' cost-contributions. The centralized DAC facilities at sequestration sites are more expanded than in the \modHybrid{} and are requiring additional heat from biomass CHPs.

>>>>

\title{Assessing \Htwo{} and \COtwo{} Network Strategies in Europe}
\title{Competition and Synergies of \Htwo{} and \COtwo{} Networks in Europe}

% \title{Synthetic fuels in Europe: Transport Hydrogen to Carbon, or Carbon to Hydrogen?}

@@

The transition to a carbon-neutral European economy is a pressing challenge that demands coordinated action across various energy sectors. While management of both \COtwo{} and \Htwo{}, also refered to as carbon and hydrogen, is considered a critical component of this transition, a gap exists in understanding how new hydrogen infrastructure effectively interacts with carbon management technologies, including of carbon capture, transport, storage, utilization, and sequestration. Hydrogen offers an efficient way to transport and store energy over long distances and timeframes, and can be produced from electrolysis or steam methane reforming (SMR), for example. Carbon can be stored in geological formations to permanently extract it from the atmosphere, a process known as carbon sequestration (CS). Additionally, carbon can be combined with hydrogen to produce carbonaceous fuels as for example needed for aviation in form of synthetic kerosene from Fischer-Tropsch synthesis, for shipping in form of synthetic methanol and for industrial feedstocks through synthetic methane. From hereon, we jointly refer to the production of these carbonaceous fuels as Carbon Utilization (CU). Carbon can be effectively captured from industrial processes and the combustion of biomass, fossil fuels, or synthetic carbonaceous fuels through point-source carbon capture (CC) techniques, or harvested from the atmosphere using direct air capture (DAC). Both \COtwo{} and \Htwo{} networks are likely to play a crucial role in the cost-effective integration of these technologies to enable net-zero economies.
% [TODO: integrate definition of Carbon Management (CM) into last sentence.]. COMMENT: CM was already mention earlier, don't repeat it instead changed framing of last sentence

% TODO: Move to better place in paper
We argue that it is crucial to consider CC, CU and CS as separate technologies, rather than grouping them under the often-used but misleading umbrella term ``carbon capture, utilization and storage'' (CCUS), to allow for independent optimization and adequate representation of each function in the carbon management system.

Recently, policymakers and industry in Europe have started developing hydrogen and carbon management strategies~\cite{GermanyDevelopingStrategy2023,CarbonManagementStrategie}, planning infrastructure components~\cite{CONetz}, and committing on the first carbon utilization projects~\cite{EFuelsPilotPlant2022,OrstedAssumesFull,GROUNDBREAKINGEFUELPRODUCTION,DLREfuelsDLR}. With the European Union's goal of achieving climate neutrality by 2050, a wide range of programs, funding models and initiatives have been established with dedicated instruments supporting both the ramp-up of the hydrogen and carbon economy~\cite{eu2023netzero,europeangreendeal,europeaninnovationfund}.
Collaborative industry initiatives like the European Hydrogen Backbone~\cite{gasforclimateEuropeanHydrogenBackbone2022} or the Hydrogen Infrastructure Map~\cite{H2InfrastructureMap} showcase the potential of hydrogen as a fuel and energy carrier. Some natural gas pipelines are already repurposed to transport hydrogen~\cite{RohrFreiFuer}. Business models from companies like Tree Energy Solutions~\cite{TESHydrogenLife2023}, Carbfix~\cite{WeTurnCO2}, and Equinor~\cite{adomaitisEquinorRWEBuild2023} advertise carbon management hubs that provide green hydrogen, methane, and synfuels on the one hand and offer \COtwo{} offtake on the other hand. The Northern Lights project in Norway~\cite{NorthernLightsWhat} is planning with a transport and sequestration capacity of 1.5~Mt \COtwo{} per year to be operative in 2024, expanding to a targeted scale of 5~Mt per year of sequestration by 2030.
The market potential for carbon capture and sequestration in Europe is underscored by the Capture Map's estimated potential~\cite{ToolsGreenTransition} of 1.7 Gt carbon annually from point sources, accounting for about half of the continent's emissions, as well as estimated sequestration capacity potentials of up to 3~Gt/a~\cite{europeancommissionEuropeanCO2Storage}. However, it's important to note that carbon sequestration technology, unlike carbon capture, is not yet fully mature and its full potential and implications remain to be explored in depth.
To advance the carbon economy, the Clean Air Task Force emphasizes the importance of developing a carbon transport system in Europe alongside a hydrogen network~\cite{lockwoodEuropeanStrategyCarbon}. Carbon pipelines, considered a mature technology, have seen widespread installations in the United States and Canada, primarily for enhanced oil recovery~\cite{righettiSitingCarbonDioxide2017,friedmannNETZEROGEOSPHERICRETURN}.

However, up to this point, it remains unclear how the two transport systems of hydrogen and carbon may complement or replace each other. Both can bridge the misalignment of sources and sinks for carbon and hydrogen which may well be determined by exogenous factors. A hydrogen network can supply regions with geographically fixed hydrogen demand, such as for steelmaking, with hydrogen from regions with the best renewable resources as well as enable CU at the site of point-source CC. On the other hand, the carbon transport system allows for transporting captured carbon to regions with sequestration potentials or high-quality renewable resources, the latter enabling cost-effective CU.
In the literature, the two network approaches and underlying technologies have been discussed in a number of publications, all of which, however, dealt with the isolated effects~\cite{bakkenLinearModelsOptimization2008,morbeeOptimisedDeploymentEuropean2012,stewartFeasibilityEuropeanwideIntegrated2014,oeiModelingCarbonCapture2014,elahiMultiperiodLeastCost2014,burandtDecarbonizingChinaEnergy2019,middletonSimCCSOpensourceTool2020,bjerketvedtOptimalDesignCost2020,weiProposedGlobalLayout2021,damoreOptimalDesignEuropean2021,becattiniCarbonDioxideCapture2022,neumannBenefitsHydrogenNetwork2022}. Such techno-economic models, in contrast for example to Integrated Assessment Models, can account for the spatial distribution of carbon sources and sinks which are crucial to provide a holistic view of the energy system and its technological interactions. A remarkable example was developed by Neumann et al.~\cite{neumannBenefitsHydrogenNetwork2022}, which comprehensively demonstrates the interchangeability of hydrogen and electricity grid expansion in a climate-neutral European energy system at the expense of higher investments. The underlying, highly-resolved model encompasses most relevant sectors, however neglects the option to transport carbon across regions.
The publication by Morbee et al.~\cite{morbeeOptimisedDeploymentEuropean2012} optimizes the topology and capacity of a \COtwo{} network in Europe, but only considers the power sector without co-optimizing renewable deployment. This limited sectoral scope cannot capture important dynamics of carbon management, since it neglects the sectors that will need to handle most \COtwo{} in the future.
A comprehensive example is found in~\cite{becattiniCarbonDioxideCapture2022}, which presents a mixed-integer model to optimize the time-evolution of a \COtwo{} transport system in Switzerland, connecting to a remote sequestration site in Norway. However, this limited spatial scope fails to consider other sequestration sites and co-benefits from connecting the \COtwo{} network to neighboring countries.
Other high-resolution energy system models do not feature detailed CU technologies and \COtwo{} transport. Euro-Calliope, for example, does neither feature carbon nor hydrogen transport, and thus forces generic synthetic fuel production to rely on captive electrolysis and DAC in the same region~\cite{pickeringDiversityOptionsEliminate2022}.

To our knowledge, no study has yet considered the co-optimization and comprehensive assessment of both \COtwo{} and \Htwo{} networks in a fully sector-coupled energy system. However, we argue that such an assessment is strongly needed to identify realms of competition and synergy between hydrogen and carbon management technologies. In this paper, we present a detailed study of the European energy system for 2050, which includes high geographical resolution and a comprehensive representation of carbon management technologies. The study is conducted using the PyPSA-Eur energy system model and encompasses all relevant energy sectors. We investigate the competitive transport dynamics between \COtwo{} and \Htwo{} networks. This is done by contrasting the following four models that feature neither one of these networks, only one, or both combined:

\begin{itemize}
    \item[] \textit{Baseline:} Neither \COtwo{} nor \Htwo{} networks are constructed.
    \item[] \textit{\COgrid{}:} Only the \COtwo{} network is developed.
    \item[] \textit{\Hgrid{}:} Only the \Htwo{} network is developed.
    \item[] \textit{Hybrid:} Both \COtwo{} and \Htwo{} networks are developed.
\end{itemize}

Anticipating with tight \COtwo{} emission targets and limited sequestration assumptions, the models are exposed to the following two scenarios:
\begin{itemize}
    \item[] \textit{Net\=/Zero}: aligning with the EU's 2050 emission targets. Limiting carbon sequestration to 200~Mt/a. Also denoted as NZ.
    \item[] \textit{Net\=/Negative}: 10\% net-negative emissions relative to 1990 levels, equaling 460~Mt \COtwo{} annually. Limiting carbon sequestration to 660~Mt/a. Also denoted as NN.
\end{itemize}

We also analyze how an energy system with limited annual sequestration potential prioritizes decarbonization and fuel switching in various sectors, and how the construction of carbon and hydrogen networks varies given net-zero or net-negative emissions targets.

In the following we show how a separate investment into no or only one of the transport technologies leads to inefficient market mechanisms. We do so, by first focusing on the Net-Zero scenario and a detailed description of the different systemic implications of the model. A analysis of the .

@@

\begin{figure}[ht!]
    \centering
    \includegraphics[width=\linewidth]{comparison/default/figures/90_nodes/cost_bar.png}
    \caption[short]{Total annual system cost subdivided into groups of technologies for the different models of the European energy system with a net-zero emission target. While the \modBase{} has neither a \COtwo{} nor an \Htwo{} network, the \modHybrid{} is allowed to expand both. ``Gas Infrastructure'' combines gas facilities for power and heat production, ``\COtwo{} Infrastructure'' and ``\Htwo{} Infrastructure'' combine transport and storage for each carrier. ``Carbon Capture at Point Sources'' combines all technologies with integrated carbon capture, including the cost of the main facility (e.g., CHP unit) and the carbon capture application.}
    \label{fig:cost_bar}
\end{figure}

Total annual system costs in the \modBase{} are highest, totalling \label{}779 billion euros. The models with additional transport options for carbon and hydrogen achieve a more efficient system allocation and lower total system costs, with \label{}3.1\% cost reduction (\label{}754~bn€) in the \modCO{}, and \label{}4.8\% (\label{}742~bn€) in the \modH{}, and \label{}5.2\% (\label{}738~bn€) in the the \modHybrid{}.

@@

% TODO: add interpretative paragraph: no CU where low-price ely, H2 is imported.
% Both networks available to smooth prices; CS draws in CC from adjacent regions; CU everywhere locally, except the CS sites; DAC centered on (very few) regions with optimal RES (and Italy/Greece for methanation)
% compare to H2 Grid model: küstennahe regionen mit longest distance to good RES(?) anzapfen, dort kein CU, CU jewtzt im Hinterland

@@

% TODO: start with explanation Fig 6: similarities b/w all models: sequestration, more DAC (=+wind, solar, heat pump, battery), point-source CC (+bioenergy); diff: level of DAC vs level of point-source CC, co2 network, h2 network

% carbon:
% - fundamentally new network layout
% - CU in spain is now supplied by carbon from DAC
% - carbon in Central Europe is mainly transported to seq sites
% - stranded investment in carbon network of 3bn

% hydrogen:
% - no significant changes in strategies

% hybrid:
% - more CU in spain
% - less cu in central europe
% - more sequestration in central europe and Portugal
% - less h2 import from spain to central european
% - mostly additive investments, except for some CU plants near the coast (of 4.3bn) and h2 network from Spain to Central Europe (2bn)

% TODO: add operation plots into appendix all models (hydrogen and carbon)

\ac{}In transitioning from a net zero to a net emissions reduction target, a continuation of investment trends are perceived. Fig.~\ref{fig:net-negative_cost_bar} shows the net investment changes per technology group for the \carbonmodel, the \Hgrid{} and \modHybrid{}s. The scenario imposes a net carbon removal of 460 Mt/a, with all models incurring a roughly similar net cost increase - 77 billion €/a for the \modCO{}, 81 billion €/a for the \modH{}, and 78 billion €/a for the \modHybrid{}. Both models invest relatively evenly in additional DAC, wind and heat pump systems, with DAC accounting for more than a third of the additional system costs. These remove carbon directly at sequestration sites at the coast, in particular in Portugal and UK, together with heat and power from new heat pumps at site and new solar and wind installation in the vicinity.

Fundamental difference in investment strategies occur in the carbon capture sector: In contrast to the \modH{}, the \modHybrid{} continues the strategy of sequestering carbon captured and transported from bio-energetic inputs, expanding the carbon transport routes from Central and Eastern Europe to the near shores. At the same time, carbonaceous fuel production in Central Europe partially moves to Spain, where it uses carbon from new DAC facilities without the installation of a \COtwo{} network. At the same time, less hydrogen imports from the Iberian Peninsula to Central Europe is needed due to less CU.
The \modH{} on the other hand, expands decentralized CU systems from the Net-Zero scenario, which are now increasingly supplied with carbon from biogas-to-gas facilities, showing up in both ``Bioenergy'' and ``Carbon Capture at Point Sources'' cost-contributions. The centralized DAC facilities at sequestration sites are more expanded than in the \modHybrid{} and are requiring additional heat from biomass CHPs.
